Dissecting the roles of local packing density and longer-range effects in protein sequence evolution

### 2017-01-19 manuscript/voronoi/dissecting-roles-local/main.tex
final article source files added

\usepackage[super,comma,sort&compress]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{fancyhdr}

% allow floats to take up as much space as they want on a page
\renewcommand{\topfraction}{1.}	% max fraction of floats at top
\renewcommand{\bottomfraction}{1.}
\renewcommand{\textfraction}{0.}
\renewcommand*{\thefootnote}{\fnsymbol{\star}}

@@

\begin{abstract}
What are the structural determinants of protein sequence evolution? A number of site-specific structural characteristics have been proposed, most of which are broadly related to either the density of contacts or the solvent accessibility of individual residues. Most importantly, there has been disagreement in the literature over the relative importance of solvent accessibility and local packing density for explaining site-specific sequence variability in proteins. We show here that this discussion has been confounded by the definition of local packing density. The most commonly used measures of local packing, such as the contact number and the weighted contact number, represent by definition the combined effects of local packing density and longer-range effects. As an alternative, we here propose a truly local measure of packing density around a single residue, based on the Voronoi cell volume. We show that the Voronoi cell volume, when calculated relative to the geometric center of amino-acid side chains, behaves nearly identically to the relative solvent accessibility, and both can explain, on average, approximately 34\% of the site-specific variation in evolutionary rate in a data set of 209 enzymes. An additional 10\% of variation can be explained by non-local effects that are captured in the weighted contact number. Consequently, evolutionary variation at a site is determined by the combined action of the immediate amino-acid neighbors of that site and of effects mediated by more distant amino acids. We conclude that instead of contrasting solvent accessibility and local packing density, future research should emphasize the relative importance of immediate contacts and longer-range effects on evolutionary variation.
\end{abstract}

@@

A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \cite{Overingtonetal1992, goldman_assessing_1998, MirnyShakhnovich1999, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross_species_2013, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number (CN) \cite{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \cite{MirnyShakhnovich1999, DokholyanShakhnovich2001, Liberlesetal2012, echave_relationship_2014, Dasmehetal2014}, protein designability \cite{england_structural_2003, Jacksonetal2013}, and measures of local flexibility, such as the B factor \cite{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014}, or flexibility measures based on elastic network models \cite{liu_sequence_2012, huang_mechanistic_2014} and Molecular Dynamics (MD) simulations \cite{shahmoradi_predicting_2014}.

It is conceivable that the majority of these site-specific quantities predicting evolutionary rate (ER) simply act as different proxy measures of an underlying biophysical constraint, such as the strength of amino acid interactions at individual sites. Indeed, Huang et al. \cite{huang_mechanistic_2014} recently proposed that the underlying constraint is mechanistic stress, which can be estimated by calculating the weighted contact number (WCN) at each site. Quantities such as relative solvent accessibility and local flexibility are expected to correlate with WCN, because all these measures depend to some extent on the density of contacts around an amino acid. Huang and coworkers thus argued that the \emph{local packing density} is the primary constraint of protein sequence evolution \cite{huang_mechanistic_2014, yeh_site-specific_2014, yeh_local_2014}.

However, a priori it is not clear whether quantities based on the contact-number concept (either CN or WCN) are truly measures of \emph{local} packing density. All these measures involve adjustable free parameters in their definitions, and depending on the parameter choices, the measures can incorporate substantial long-range effects. In particular, WCN by default uses an inverse-square weighting, which assumes that amino acids at an arbitrary distance away from the focal residue still exert some measurable effect on that residue. Thus, some of the best predictors of evolutionary variation seem to confound local and longer-range effects, and the relative importance of the immediate neighborhood of an amino acid vs.\ longer range effects is not known.

@@

We analyzed the same data set of 209 monomeric enzymes we have previously analyzed \cite{echave_relationship_2014} and that was originally published with four additional proteins \cite{yeh_site-specific_2014}. The enzymes were originally randomly picked from the Catalytic Site Atlas $2.2.11$ \cite{porter_catalytic_2004}, range from 95 to 1287 amino acids in length, and include representatives from all six main EC functional classes \cite{webb_enzyme_1992} and domains of all main SCOP structural classes \cite{murzin_scop:_1995}. Evolutionary rates at all sites in all proteins were calculated as described \cite{echave_relationship_2014}. In brief, we employed the software Rate4Site \cite{mayrose_comparison_2004}, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model.

@@

We used the DSSP software \cite{kabsch_dictionary_1983} to calculate  the Accessible Surface Area (ASA) for each residue in each protein, and we normalized these ASA values by their theoretical maximum \cite{tien_maximum_2013} to obtain Relative Solvent Accessibility (RSA).

@@

We used the VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset. Among the most important properties are the length of the cell edges, surface area and volume, number of faces of each cell, and the cell eccentricity, defined as the distance between the cell's seed and the geometrical center of the cell. In addition, the cell {\it sphericity}, which is a measure of the compactness of the cell, can be defined as
    \begin{equation}
        \label{eqn:sphericity}
        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
    \end{equation}
Here, $V$ and $A$ represent the volume and the area of the cell, respectively.  By definition, $\Psi$ falls between 0 and 1. The limit of $\Psi=1$ indicates that the cell is perfectly spherical, while the limit of $\Psi=0$ indicates that the cell is a 2-dimensional object that has no volume but only surface area.

@@

    One potential caveat with Voronoi tessellation of finite structures is the existence of {\it edge effects}. Sites that are close to the surface of the protein can be associated with Voronoi cells that are open, only bounded by the cubic box containing the protein. We performed the following analysis to verify that these edge effects did not influence the observed sequence-structure correlations. We first identified all open cells in all proteins by examining the fraction of change in the cell volumes upon changing the size of the cubic box that contained the protein. For each protein, once the open cells were identified, we multiplied their volumes by a normalizing factor such that the smallest-volume open cell had a larger volume than largest closed cell. We then ordered the open cells by the fraction of volume change observed upon changing the box size. In other words, the more the cell volume increased upon increasing the box size, the larger the volume of that cell was assumed to be, compared to all other open cells in the protein. Finally, we calculated Spearman's correlation coefficient between this ordered set of cell volumes (that included both open and closed cells) and the sequence evolutionary rates. We then compared the computed correlation coefficients to those obtained by using the original cell volumes without corrections for edge effects. The resulting distributions of correlation coefficients for the original and corrected cell volumes were nearly identical (mean difference $<0.001$).

For our dataset of 209 monomeric proteins, we concluded that edge effects due to Voronoi tessellation appear to have virtually no influence on the observed sequence-structure correlations. We reached similar conclusions when open cells were alternatively ranked by other criteria, such as the fractional changes in cell area upon changing the box size. Thus, the Voronoi cell characteristics, in particular cell volume and cell area, can be safely used in predicting sequence variability, as long as a sufficiently large enclosing box is used to calculate the Voronoi cell characteristics of open cells.

@@

@@

Prior work on the same enzyme data set that we have analyzed here has shown that contact number generally outperforms all other structural predictors of site variability \cite{yeh_site-specific_2014, yeh_local_2014, huang_mechanistic_2014}. This finding has been interpreted as indicating that local packing density is the primary evolutionary constraint imposed by protein structure. However, to what extent the contact number (CN) is actually a truly local measure is unclear. The contact number at a site $i$, CN$_i$, is defined as the number of amino acids within a fixed radius $r_0$ from that site,
    \begin{equation}
        \label{eqn:cn}
        \text{CN}_{i} = \sum_{j\neq i} H\big(r_0-r_{ij}\big),
    \end{equation}
where $r_{ij}$ is the distance between sites $i$ and $j$ and $H(r)$ is the Heaviside step function
    \begin{equation}
        \label{eqn:heaviside}
        H(r) = \begin{cases}
        			1 \quad\text{for $r\leq 0$,}\\
                    0 \quad\text{otherwise.}
               \end{cases}
    \end{equation}
The location of individual sites is commonly chosen as the position of the residues' C$_\alpha$ backbone atoms, but recent work has shown that the geometric center of the side chain may be a better choice \cite{marcos_too_2015}. Irrespective of the particular choice of reference point for each site, how local of a measure CN$_i$ actually is depends on the arbitrary cutoff parameter $r_0$. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range of 5\AA\ to 18\AA\ \cite{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014, yeh_local_2014}. While a cutoff of 5\AA\ likely captures only immediate neighbors of the focal amino acid, a cutoff of 18\AA\ will capture many amino acids that do not directly interact with the focal amino acid, and thus captures both local and non-local packing-density information.

@@

Evidently, both CN and WCN perform best at describing ER variation when they capture a non-negligible contribution of longer-range effects, beyond the immediate neighborhood of each focal site. To disentangle the contributions of local and longer-range effects to sequence evolution, we next proceeded to develop a measure of packing density that by definition depended on only the immediate neighbors around a single amino acid.

@@

In this context, we would like to emphasize that most Voronoi cell characteristics increase with decreasing local packing density. For example, the larger the cell volume or cell surface area, the fewer amino acids are located in the immediate neighborhood of the focal residue. By contrast, increasing sphericity implies increasing packing density. In general, a polyhedron with a higher number of faces corresponds to higher sphericity, the limiting case of which is a perfect sphere with an infinite number of faces. In the context of Voronoi polyhedra in proteins, a higher number of faces of a cell indicates more amino-acid neighbors around the site of interest, and therefore more interactions among them. Consequently, correlations with cell sphericity tend to have the opposite sign from correlations with other Voronoi cell characteristics.

Not unexpectedly, we found that the Voronoi cell characteristics correlated strongly with each other. Therefore, we next used cell volume (measuring inverse local packing density) as the reference and calculated the partial correlations of ER with Voronoi cell properties controlling for cell volume. We found that these partial correlations were generally quite low, with median absolute values of approximately 0.1  (Figure \ref{fig:voronoi_ER_screen}B). Cell edge length and cell sphericity displayed negative partial correlations with ER while cell eccentricity and cell area displayed positive partial correlations. In conclusion, with regards to ER variation, the cell area, volume, and edge length appear to represent largely the same underlying property of a Voronoi cell, its size.

@@

We found that side-chain WCN displayed substantial residual correlations with ER after controlling for Voronoi cell volume (Figure \ref{fig:best_predictorER}B). The distribution of (absolute) partial correlation coefficients had a median value of 0.32, with an interquartile range of 0.17 (from 0.23 to 0.40). Thus, after controlling for local packing, the longer-range effects appeared to explain approximately $10\%$ of the site-specific sequence evolutionary rates. Similarly, there were significant residual correlations of C$_\alpha$ WCN with ER after controlling for Voronoi cell volume, with a median (absolute) correlation coefficient of $0.21$. By contrast, the residual correlations of RSA with ER were by-and-large negligible. The median was only $0.09$, with an interquartile range of 0.17 (from $-0.02$ to $0.15$).

Thus, in conclusion, either WCN definition (side-chain or C$_\alpha$) provided a significant, independent contribution to explaining ER when purely local effects of packing were controlled for by Voronoi cell volume. By contrast, RSA did not make an independent contribution once Voronoi cell volume was controlled for, as was expected given these quantities' strong correlations with each other.

@@

In this work, we have shown that the commonly used measures of packing density, contact density (CN) and weighted contact density (WCN), capture non-local effects of amino-acid packing in a protein. To quantify the magnitude of these effects, we have defined purely local measures of packing density via the Voronoi tessellation, and we have shown that Voronoi cell volume encapsulates most of the relevant information about packing contained in various Voronoi cell characteristics. We have then shown that the variation in site-specific evolutionary rates (ER) can be partitioned into two independent components: variation explained by local, site-specific packing density (as measured by Voronoi cell volume) and variation explained by more long-range effects that extend beyond the first coordination shell around each site in a given protein. The longer-range effects play a non-negligible role in protein sequence evolution and explain on average approximately 10\% of the observed variability in ER in the data set studied here. These 10\% are in addition to and independent of the proportion of variance explained by the site-specific packing density, which explains on average another 34\%. By contrast, the relative solvent accessibility (RSA) has negligible explanatory power for ER once site-specific packing is taken into account via Voronoi cell volume.

@@

@@

@@

One limitation of our present study is the choice of data set, 209 monomeric enzymes that have been previously studied by multiple groups. All enzymes in this data set are relatively small, globular, and have a well-defined active site, usually close to the center of the protein \cite{shih_evolutionary_2012}. It is possible that the weighted contact number, with its radial weighting, performs particularly well on such proteins. For example, in the large and elongated influenza protein hemagglutinin, RSA explains over twice the variation in evolutionary rate than WCN does  \cite{meyer_geometric_2015}. Moreover, active sites and protein--protein interfaces generally induce additional evolutionary constraints, and how these constraints interact with local packing density and longer-range steric interactions remains largely unexplored \cite{Deanetal2002, franzosa_structural_2009, meyer_geometric_2015,Abriataetal2015}. Therefore, future work will have to disentangle the contributions of local packing and longer-range effects in a wider set of protein structures, and it will also have to disentangle these contributions from the evolutionary constraints imposed by active sites and protein--protein interfaces.

\section*{Appendix A: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
\label{app:best_crd}

>>>>

\usepackage[super,comma,sort&compress]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{fancyhdr}
\usepackage{setspace}
\doublespacing

% allow floats to take up as much space as they want on a page
\renewcommand{\topfraction}{1.}	% max fraction of floats at top
\renewcommand{\bottomfraction}{1.}
\renewcommand{\textfraction}{0.}
\renewcommand*{\thefootnote}{\fnsymbol{\star}}
\newcommand{\angstrom}{\text{\normalfont\AA}}

@@

\begin{abstract}
What are the structural determinants of protein sequence evolution? A number of site-specific structural characteristics have been proposed, most of which are broadly related to either the density of contacts or the solvent accessibility of individual residues. Most importantly, there has been disagreement in the literature over the relative importance of solvent accessibility and local packing density for explaining site-specific sequence variability in proteins. We show that this discussion has been confounded by the definition of local packing density. The most commonly used measures of local packing, such as contact number and the weighted contact number, represent the combined effects of local packing density and longer-range effects. As an alternative, we propose a truly local measure of packing density around a single residue, based on the Voronoi cell volume. We show that the Voronoi cell volume, when calculated relative to the geometric center of amino-acid side chains, behaves nearly identically to the relative solvent accessibility, and each individually can explain, on average, approximately 34\% of the site-specific variation in evolutionary rate in a data set of 209 enzymes. An additional 10\% of variation can be explained by non-local effects that are captured in the weighted contact number. Consequently, evolutionary variation at a site is determined by the combined effects of the immediate amino-acid neighbors of that site and effects mediated by more distant amino acids. We conclude that instead of contrasting solvent accessibility and local packing density, future research should emphasize on the relative importance of immediate contacts and longer-range effects on evolutionary variation.
\end{abstract}

@@

A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \cite{Overingtonetal1992, goldman_assessing_1998, MirnyShakhnovich1999, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross_species_2013, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number (CN) \cite{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \cite{MirnyShakhnovich1999, DokholyanShakhnovich2001, Liberlesetal2012, echave_relationship_2014, Dasmehetal2014}, protein designability \cite{england_structural_2003, Jacksonetal2013}, and measures of local flexibility, such as the B factor \cite{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014}, or flexibility measures based on elastic network models \cite{liu_sequence_2012, huang_mechanistic_2014} and Molecular Dynamics (MD) simulations \cite{shahmoradi_predicting_2014}, or direct comparison of fluctuations among static structures of homologous proteins \cite{fuglebakk_measuring_2012,shahmoradi_predicting_2014, carvalho_comparison_2015}. (For a recent review, see Ref.~\cite{echave_causes_2016}.) It is conceivable that the majority of these site-specific quantities predicting evolutionary rate (ER) simply act as different proxy measures of an underlying biophysical constraint, such as the strength of amino acid interactions at individual sites. Indeed, Huang et al. \cite{huang_mechanistic_2014} recently proposed that the underlying constraint is mechanistic stress, which can be estimated by calculating the weighted contact number (WCN) at each site. Quantities such as relative solvent accessibility and local flexibility are expected to correlate with WCN, because all these measures depend to some extent on the density of contacts around an amino acid\cite{echave_causes_2016}. Huang and coworkers thus argued that the \emph{local packing density} is the primary constraint of protein sequence evolution \cite{huang_mechanistic_2014, yeh_site-specific_2014, yeh_local_2014}.

However, a priori it is not clear whether quantities based on the contact-number concept (either CN or WCN) are truly measures of \emph{local} packing density. All these measures involve adjustable free parameters in their definitions, and depending on the parameter choices, the measures can incorporate substantial long-range effects. In particular, WCN by default uses an inverse-square weighting, which assumes that amino acids at an arbitrary distance away from the focal residue still exert some measurable effect on that residue. Thus, some of the best predictors of evolutionary variation seem to confound local and longer-range effects, and the relative importance of the immediate neighborhood of an amino acid vs.\ longer range effects is not known. Importantly, by longer-range effects we mean here effects between amino acids that are distant in 3D space. Amino acids that are distant in sequence but close in 3D space, as are frequently observed\cite{plaxco_contact_1998,lai_sequence_2015}, would contribute to local packing but not to longer-range effects by our definition.

@@

We analyzed the same data set of 209 monomeric enzymes we have previously analyzed \cite{echave_relationship_2014} and that was originally published with four additional proteins \cite{yeh_site-specific_2014}. The enzymes were originally randomly picked from the Catalytic Site Atlas $2.2.11$ \cite{porter_catalytic_2004}, range from 95 to 1287 amino acids in length, and include representatives from all six main EC functional classes \cite{webb_enzyme_1992} and domains of all main SCOP structural classes \cite{murzin_scop:_1995}. Evolutionary rates at all sites in all proteins were calculated as described \cite{echave_relationship_2014}. In brief, we employed the software Rate4Site \cite{mayrose_comparison_2004}, using the empirical Bayesian method and the amino-acid Jukes-Cantor-like (JC-like) mutational model.

@@

We used the DSSP software \cite{kabsch_dictionary_1983} to calculate  the Accessible Surface Area (ASA) for each residue in each protein, and we normalized these ASA values by their theoretical maximum \cite{tien_maximum_2013} to obtain Relative Solvent Accessibility (RSA). Throughout this work, we considered RSA rather than ASA, consistent with common practice in the literature. RSA is preferred in evolutionary analyses because ASA confounds residue size and residue exposure. The maximum ASA varies by a factor of three from the smallest amino acid (glycine) to the largest (tryptophan) \cite{tien_maximum_2013}. As a consequence, a fully exposed glycine may have the same ASA as a largely but not completely buried tryptophan. By contrast, an RSA of 1 indicates that the residue is fully exposed, regardless of the nature of the residue. Thus, we generally expect RSA to be less variable under amino-acid substitution than ASA. Fully exposed residues tend to remain fully exposed, partially exposed residues tend to remain partially exposed, and so on.

@@

We used the VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset. Specifically, we calculated the length of the cell edges, surface area and volume, number of faces of each cell, and the cell eccentricity, defined as the distance between the cell's seed and the geometrical center of the cell. We also calculated the cell {\it sphericity}, which is a measure of the compactness of the cell, defined as
    \begin{equation}
        \label{eqn:sphericity}
        \Psi = \frac{\pi^{\frac{1}{3}}(6V)^{\frac{2}{3}}}{A}.
    \end{equation}
Here, $V$ and $A$ represent the volume and the area of the cell, respectively.  By definition, $\Psi$ falls between 0 and 1. The limit of $\Psi=1$ indicates that the cell is perfectly spherical, while the limit of $\Psi=0$ indicates that the cell is a 2-dimensional object that has no volume but only surface area.

@@

When performing a Voronoi tessellation of a protein, which has a finite size, we have to be aware of the existence of {\it edge effects} on the protein surface. Most Voronoi cells on the protein surface are not completely enclosed by other neighboring cells. Therefore, to ensure that all cells are closed in all directions, we need to place the protein structure into a hypothetical box of arbitrary size. This box provides the missing boundaries for the otherwise open cells on the protein surface. Voronoi cells that are bounded by this box tend to have artifactual properties, such as much larger volumes and areas compared to the cells that are fully surrounded by other neighboring cells. In addition, the volumes and areas of these cells depend on the size of the box, which is arbitrary.

To verify that these edge effects did not influence the observed sequence-structure correlations, we investigated how our results changed as a function of the size of the enclosing box. To do so, we first identified all open cells in all proteins by examining the fraction of change in the cell volumes upon changing the size of the containing box.  Initially, each wall of the box was set to minimum distance of $30.0$\AA\ from all atoms in the protein, and then the size of the box was changed, such that each wall would be at least $60.0$\AA\ away from any atom in the protein.   Note that changing the size of the box surrounding the protein will only affect the properties of those cells that are inherently open in the absence of the box.

For each protein, once open cells were identified, we sorted these cells according to their rates of volume change upon changing the box size, such that the slowest changing cell was assumed to have the smallest volume among open cells, and the fastest changing cell was assumed to have the largest volume among open cells. Note that this method of ranking the open cells is very similar to ranking the cells by the solid opening angle of open cells, without the requirement of explicitly measuring the solid opening angles for all open cells. The underlying assumption here is that cells with larger solid opening angles will show larger rate of cell volume change upon changing the box size surrounding the protein. The wider the solid opening angle of a cell is, the more space is assumed to be available to the amino acid residing the open cell. In essence, the rate of volume change provides a method of ranking the open cells based on their solid opening angles, without direct calculation of the angles.

All open cells were assumed to have a larger volume than any closed cell.  Finally, we calculated Spearman's correlation coefficient between this ordered set of cell volumes (that included both open and closed cells) and the evolutionary rates. We then compared the computed correlation coefficients to those obtained by using the original cell volumes without corrections for edge effects. The resulting distributions of correlation coefficients for the original and corrected cell volumes were nearly identical (mean difference $<0.001$).

For our dataset of 209 monomeric proteins, we thus concluded that edge effects due to Voronoi tessellation appear to have virtually no influence on the observed sequence-structure correlations. We reached similar conclusions when open cells were alternatively ranked by other criteria, such as the fractional changes in cell area upon changing the box size. Thus, the Voronoi cell characteristics, in particular cell volume and cell area, can be safely used in predicting sequence variability, as long as a sufficiently large enclosing box is used to calculate the Voronoi cell characteristics of open cells.

@@

Throughout this study, we ask to what extent various measures of local packing in proteins predict evolutionary variation. To address this question, we carry out a number of analyses that all consist of the same conceptual steps illustrated in Figure \ref{fig:pipeline}. We first correlate the evolutionary rate (ER) at each site with a measure of packing at the same site (Figure \ref{fig:pipeline}A, B). We then calculate the corresponding correlation coefficients separately for all proteins in our data set (Figure \ref{fig:pipeline}C). Finally, we compare the distributions of these correlation coefficients for different measures of packing (Figure \ref{fig:pipeline}D).

@@

Prior work on the same enzyme data set that we have analyzed here has shown that contact number generally outperforms all other structural predictors of site variability \cite{yeh_site-specific_2014, yeh_local_2014, huang_mechanistic_2014}. This finding has been interpreted as indicating that local packing density is the primary evolutionary constraint imposed by protein structure. However, to what extent the contact number (CN) is actually a truly local measure is unclear. The contact number at a site $i$, CN$_i$, is defined as the number of amino acids within a fixed radius $r_0$ from that site,
    \begin{equation}
        \label{eqn:cn}
        \text{CN}_{i} = \sum_{j\neq i} H\big(r_0-r_{ij}\big),
    \end{equation}
where $r_{ij}$ is the distance between sites $i$ and $j$ and $H(r)$ is the Heaviside step function
    \begin{equation}
        \label{eqn:heaviside}
        H(r) = \begin{cases}
        			1 \quad\text{for $r\geq 0$,}\\
                    0 \quad\text{otherwise.}
               \end{cases}
    \end{equation}
The location of individual sites is commonly chosen as the position of the residues' C$_\alpha$ backbone atoms, but recent work has shown that the geometric center of the side chain may be a better choice \cite{marcos_too_2015}. Irrespective of the particular choice of reference point for each site, how local of a measure CN$_i$ actually is depends on the arbitrary cutoff parameter $r_0$. There is no consensus on the optimal value of this cutoff distance, although it is typically chosen in the range of 5\AA\ to 18\AA\ \cite{lin_deriving_2008, franzosa_structural_2009, weng_molecular_2014, yeh_local_2014}. While a cutoff of 5\AA\ likely captures only immediate neighbors of the focal amino acid, a cutoff of 18\AA\ will capture many amino acids that do not directly interact with the focal amino acid, and thus captures both local and non-local packing-density information.

@@

Next we asked whether the improved performance of WCN over CN was caused specifically by the power-law weighting of close contacts or by the inclusion of long-range interactions. To do so, we considered a hybrid WCN model with power-law weighting and a hard cutoff beyond which all other residues were excluded. More specifically, we fixed the exponent of the power-law weighting function to $\alpha=-2$ and further included only those sites in the calculation of WCN that were within a specific radius of neighborhood $r_0$ from the site of interest,
    \begin{equation}
        \label{eqn:wcn_pwrl_cutoff}
        \text{WCN}_{i}^\text{cutoff} = \sum^N_{j\neq i} r_{ij}^{-2} H\big(r_0-r_{ij}\big),
    \end{equation}
\noindent with $H$ defined by Eqn. \ref{eqn:heaviside}.
We then varied the cutoff $r_0$ and searched for the value that would result in the strongest correlation of WCN with site-specific evolutionary rates, separately for each individual protein in our dataset. Finally, we measured the mean fraction of neighboring sites that were required in the calculation of $\text{WCN}^\text{cutoff}$  to obtain the highest correlation strengths. We found for the vast majority of proteins that the highest correlation strengths were obtained when over 90\% of the sites in the protein were included in the calculation of $\text{WCN}^\text{cutoff}$ (Figure \ref{fig:wcn_pwrl_cutoff}). This large fraction is a clear indication of the importance of distant sites in their evolution.

In combination, these results demonstrate that both CN and WCN perform best at describing ER variation when they capture a non-negligible contribution of longer-range effects, beyond the immediate neighborhood of each focal site. To disentangle the contributions of local and longer-range effects to sequence evolution, we next proceeded to develop a measure of packing density that by definition depended on only the immediate neighbors around a single amino acid.

@@

In this context, we emphasize that most Voronoi cell characteristics increase with decreasing local packing density. For example, the larger the cell volume or cell surface area, the fewer amino acids are located in the immediate neighborhood of the focal residue. By contrast, increasing sphericity implies increasing packing density. In general, a polyhedron with a higher number of faces corresponds to higher sphericity, the limiting case of which is a perfect sphere with an infinite number of faces. In the context of Voronoi polyhedra in proteins, a higher number of faces of a cell indicates more amino-acid neighbors around the site of interest, and therefore more interactions among them. Consequently, correlations with cell sphericity tend to have the opposite sign from correlations with other Voronoi cell characteristics.

Not unexpectedly, we found that the Voronoi cell characteristics correlated strongly with each other. Therefore, we next used cell volume (measuring inverse local packing density) as the reference and calculated the partial correlations of ER with Voronoi cell properties controlling for cell volume. We found that these partial correlations were generally quite low, with median absolute values of approximately 0.1  (Figure \ref{fig:voronoi_ER_screen}B). Cell edge length and cell sphericity displayed negative partial correlations with ER while cell eccentricity and cell area displayed positive partial correlations.

@@

We found that side-chain WCN displayed substantial residual correlations with ER after controlling for Voronoi cell volume (Figure \ref{fig:best_predictorER}B, Table \ref{tab:explained_var}). The distribution of (absolute) partial correlation coefficients had a median value of 0.32, with an interquartile range of 0.17 (from 0.23 to 0.40). Thus, after controlling for local packing, the longer-range effects appeared to explain approximately $10\%$ of the site-specific sequence evolutionary rates. Similarly, there were significant residual correlations of C$_\alpha$ WCN with ER after controlling for Voronoi cell volume, with a median (absolute) correlation coefficient of $0.21$. By contrast, the residual correlations of RSA with ER were by-and-large negligible. The median was only $0.09$, with an interquartile range of 0.17 (from $-0.02$ to $0.15$).

In conclusion, either WCN definition (side-chain or C$_\alpha$) provided a significant, independent contribution to explaining ER when purely local effects of packing were controlled for by Voronoi cell volume. By contrast, RSA did not make an independent contribution once Voronoi cell volume was controlled for, as was expected given these quantities' strong correlations with each other.

\subsection*{Sensitivity of results to evolutionary-rate model matrix}

The evolutionary rates we have used throughout this work were calculated based on a Jukes-Cantor-like (JC-like) model, in which all amino-acid substitutions are assumed to be equally likely and no corrections are made for mutation biases and/or the structure of the genetic code \cite{echave_relationship_2014}. More sophisticated evolutionary models, such as the JTT model\cite{jones_rapid_1992}, the WAG model\cite{WhelanGoldman2001}, or the LG model\cite{LeGascuel2008}, attempt to correct for these factors, and generally yield improved model fit compared to simpler models. To what extent these various models affect rate inferences is not well understood. In our analysis here, we used the same rates that had been used in previous analyses \cite{echave_relationship_2014,marcos_too_2015}, so that our results were directly comparable to these previous results. However, the same enzyme data set we analyzed here has also been studied previously with rates calculated under the JTT model\cite{yeh_site-specific_2014,huang_mechanistic_2014,yeh_local_2014}, so we asked to what extent our conclusions differed among these two sets of evolutionary rates.

We found that, first, for each individual protein in the dataset, the evolutionary rates calculated under the JTT model correlated strongly with the evolutionary rates calculated under the JC-like model. The average Spearman correlation coefficient was $0.97$ with the standard deviation of distribution $\sigma=0.02$ over the entire dataset. Second, the results that we report in Table \ref{tab:explained_var} (percent variance explained by Voronoi cell volume and by WCN controlling for Voronoi cell volume) change only by approximately 2 percentage points when JTT rates are used in place of JC-like rates. Similarly, all other analyses also showed only minor variations upon using JTT rates. Therefore, we conclude that our findings are not strongly affected by the specific model with which evolutionary rates are calculated.

@@

We have shown that the commonly used measures of packing density, contact density (CN) and weighted contact density (WCN), capture non-local effects of amino-acid packing in a protein. To quantify the magnitude of these effects, we have defined purely local measures of packing density via the Voronoi tessellation, and we have shown that Voronoi cell volume encapsulates most of the relevant information about packing contained in various Voronoi cell characteristics. We have then shown that the variation in site-specific evolutionary rates (ER) can be partitioned into two independent components: variation explained by local, site-specific packing density (as measured by Voronoi cell volume) and variation explained by more long-range effects that extend beyond the first coordination shell around each site in a given protein. The longer-range effects play a non-negligible role in protein sequence evolution and explain on average approximately 10\% of the observed variability in ER in the data set studied here. These 10\% are in addition to and independent of the proportion of variance explained by the site-specific packing density, which explains on average another 34\%. By contrast, the relative solvent accessibility (RSA) has negligible explanatory power for ER once site-specific packing is taken into account via Voronoi cell volume.

One limitation of our study is the choice of the data set, 209 monomeric enzymes that have been previously studied by multiple groups. All enzymes in this data set are relatively small, globular, and have a well-defined active site, usually close to the center of the protein \cite{shih_evolutionary_2012}. It is possible that the weighted contact number, with its radial weighting, performs particularly well on such proteins. For example, in the large and elongated influenza protein hemagglutinin, RSA explains over twice the variation in evolutionary rate than WCN does  \cite{meyer_geometric_2015}. Moreover, active sites and protein--protein interfaces generally induce additional evolutionary constraints, and how these constraints interact with local packing density and longer-range steric interactions remains largely unexplored \cite{Deanetal2002, franzosa_structural_2009, meyer_geometric_2015,Abriataetal2015}. Therefore, future work will have to disentangle the contributions of local packing and longer-range effects in a wider set of protein structures, and it will also have to disentangle these contributions from the evolutionary constraints imposed by active sites and protein--protein interfaces.

@@

When searching for both the optimal cutoff for CN and the optimal exponent for WCN, we found a single global optimum in either case (Figure \ref{fig:cnwcnp}). This result was not unexpected. First, the weighting functions (Heaviside step function for CN, power-law for WCN) are monotonically decreasing functions of pairwise distance. Second, even though the pairwise interaction strengths of amino acids in a protein are complex functions of inter-residue distance, virtually unique to each pair of residues in each protein, the overall ``averaged over protein-medium'' behavior of interaction strength vs.\ inter-residue distance decays monotonically with distance \cite{dehouck_effective_2013,xia_correlation_2015,xia_multiscale_2015}. Therefore, the observed behavior in Figure \ref{fig:cnwcnp} merely assert that there is a single optimal parameter value that represents the best relative weighting of local and longer-range interactions. Either increasing or decreasing this parameter value will yield a sub-optimal relative weighting of local and longer-range interactions.

@@

Other residue--residue interactions besides steric interactions are less likely to cause the long-range effects we have observed here. Covalent bonds and van der Waals' interactions hardly reach beyond the first coordination shells of amino acids, limiting the effective ranges of these interactions to the amino acid's immediate neighborhood $\lesssim10$\AA. And disulfide bonds and ionic interactions are infrequently observed in our sample of enzymatic proteins.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%    
\section*{Appendix A: Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
\label{app:best_crd}
